Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-10-02 vertical-thermal-review.Rmd
ref updates

Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.
Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. 
The net result of higher temperatures may be increased or decreased tree growth and forest carbon sequestration, with decreases being more commonly documented across the world's forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021]. 
The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

@@

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021].
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

@@

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From @anderson-teixeiraJointEffectsClimate2021. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020. ](tree_rings/Figure_4_tree_rings.png)

@@

As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves [@ipccClimateChangeWidespread2021].
These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
Our in-depth review sheds some light as to how responses are likely to vary across forest vertical gradients, yet important uncertainties remain.

@@

As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChangeWidespread2021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011]. 
Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020]. 
While wetter forests can prove quite resilient to such changes [@millerReducedImpactLogging2011], this shift to hotter and drier microclimates can often make forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018]. 
Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].

@@

Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, as we present in this paper). 
At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @depurySimpleScalingPhotosynthesis1997; @bonanMovingIncorrectUseful2021].
In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs. `r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig. `r `fig_leaf_T`) to vertically varying abiotic conditions (Fig. `fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.

>>>>

Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.
Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. 
The net result of higher temperatures may be increased or decreased tree growth and forest carbon sequestration, with decreases being more commonly documented across the world's forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021]. 
The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

@@

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress].
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

@@

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From @anderson-teixeiraJointEffectsClimateinpress. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)

@@

As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves [@ipccClimateChange20212021].
These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
Our in-depth review sheds some light as to how responses are likely to vary across forest vertical gradients, yet important uncertainties remain.

@@

As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011]. 
Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020]. 
While wetter forests can prove quite resilient to such changes [@millerReducedImpactLogging2011], this shift to hotter and drier microclimates can often make forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018]. 
Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].

@@

Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, as we present in this paper). 
At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @depurySimpleScalingPhotosynthesis1997; @bonanMovingIncorrectUseful2021].
In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.`r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.`r `fig_leaf_T`) to vertically varying abiotic conditions (Fig.`fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.
